\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2019-06-21 IPANEMAP-2019.tex

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al} dataset, further described. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics used by previous works~\cite{Spasic2017}.
\bsi{pourquoi utilise-t-on ici la Gm des bp distance et ensuite le MCC? à expliquer?}\ypi{Pour des raisons historiques, surtout dans le benchmark suivant, j'ai rajouté une phrase dans cette direction}As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data.

@@

%Remarques These :
%\begin{itemize}
%  \item Fabrice : Y a t il une raison pour laquelle on est les meilleurs ?
%\end{itemize}
This observed discrepancy is somewhat surprising, given that \OurTool and \Software{RSample} share the same sampling/clustering based modeling. However, the two methods differ substantially whenever a single conformation is detected, in which case \Software{RSample} switches to a simple MEA prediction, whereas \OurTool{} always proceeds with an iterative clustering. This allows \OurTool{} to consider a larger number of clusters, which we believe helps eliminating structural outliers, leading to the election of a better final centroid structure.

@@

	\caption{Model of the Lariat capping ribozyme from {\itshape Didymium iridis} (A -- Meyer\etal\cite{Meyer2014}), pseudoknot-free secondary structure as annotated by \Software{DSSR} (B -- pseudoknot ommitted) and \OurTool first-ranking prediction (C.) from 8 maximally-diverging probing conditions (NMIAMg, 1M7ILU3Mg, NAIMg, 1M7, CMCTMg,1M7ILU3,1M7Mg). \bsi{figure à refaite. Faut il montrer les interactions tertiaires? } \label{fig:didy}}
\end{figure*}
As benchmark RNA we used the lariat capping ribozyme from \textit{Didymium Iridis} (DiLCrz) which 3D structure has been modeled at 3,5 A from X-ray diffraction data~\cite{Meyer2014}. This RNA contains a large diversity of interactions, including helical domains, tertiary pairings, a kissing loop and a pseudoknot (Figure~\ref{fig:didy}).
It was probed as described in material and methods in 16 different conditions using different SHAPE and classical reagents, in presence/absence of Mg$^\text{2+}$ ions, using either the standard premature stop or the newly developed mutation induction technologies to map the modification sites. These experimental results were then used as input of \OurTool{}, either individually or combined,  for secondary structure prediction.

@@

A visual inspection of Figure~\ref{fig:PCA} suggests the presence of 8 clusters. In order to objectively build groups of compatible conditions, we performed a k-mean clustering using \Software{scikit-learn}, setting k to 8, and obtained the clusters highlighted in Figures~\ref{fig:biclustering} and \ref{fig:PCA}.  We obtain the following clusters:
\begin{itemize}\setlength{\itemsep}{0pt}\setlength{\parskip}{.1em}	
\item \BullLab{A}: \OneMSevILU, \NMIA, and \NMIAMg;
\item \BullLab{B}: \OneMSevCE, \NMIACE, \BzCN, and \NAICE;
\item \BullLab{C}: \OneMSevILUThreeMg and \OneMSevILUMg;
\item \BullLab{D}: \OneMSevILUThree;
\item \BullLab{E}: \NMIAMgCE and \OneMSevMgCE;
\item \BullLab{F}: \DMSMg and probing-free;
\item \BullLab{G}: \CMCTMg and \NAIMg;
\item \BullLab{H}: \BzCNMg.
\end{itemize}
The predicted clusters are generally consistent with the ordering resulting from a spectral biclustering, implemented in \Software{scikit-learn} and executed with default parameters, as illustrated by Figure~\ref{fig:glycine_example}.

@@

\begin{table}
	\Heatset{min=60,   
		max=83,   
		max colour=Aquamarine3, % colour at maximum
		min colour=BadCol,      % colour at minimum
		Min colour=BadCol, % colour for values below min
		Max colour=Aquamarine3   % colour for values above max
	}
	\newcommand{\Mut}{Mut}
	\newcommand{\Stop}{Stop}
	\newcommand{\Mg}{$\bullet$}
	\newcommand{\NoMg}{$\circ$}

	{\centering
			\begin{adjustbox}{max width=\linewidth}
		\begin{tabular}{@{}lcccHHH}
			\toprule
			                            &          &         &           & \multicolumn{3}{@{}c@{}}{MCC\%}                                                  \\
			Condition                   &  Clust.  &  Tech.  & Mg$^{2+}$ & \multicolumn1l{\relsize{-1}\OurTool} & \multicolumn1l{MFE} & \multicolumn1l{MEA} \\ \midrule
			\OneMSevILUMg                    & \Bull{C} &  \Mut   &    \Mg    & 85                                   & 82                  & 82                \\
			\OneMSevILU                      & \Bull{A} &  \Mut   &   \NoMg   & 84                                   & 82                  & 84                \\
			\NMIAMg                      & \Bull{A} &  \Mut   &    \Mg    & 81                                   & 80                  & 80                \\
			\NMIA                        & \Bull{A} &  \Mut   &   \NoMg   & 80                                   & 69                  & 69                \\
			\OneMSevILUThreeMg                   & \Bull{C} &  \Mut   &    \Mg    & 74                                   & 75                  & 75                \\
			\NMIAMgCE                    & \Bull{E} &  \Stop  &    \Mg    & 73                                   & 72                  & 73                \\
			\NAIMg                       & \Bull{G} &  \Stop  &    \Mg    & 73                                   & 68                  & 70                \\
			\BzCNMg                      & \Bull{H} &  \Stop  &    \Mg    & 71                                   & 68                  & 54                \\
			\OneMSevMgCE                     & \Bull{E} &  \Stop  &    \Mg    & 70                                   & 69                  & 70                \\
			\CMCTMg                      & \Bull{G} &  \Stop  &    \Mg    & 70                                   & 68                  & 73                \\
			\OneMSevILUThree                     & \Bull{D} &  \Mut   &   \NoMg   & 64                                   & 56                  & 61                \\
			\DMSMg                       & \Bull{F} &  \Stop  &    \Mg    & 63                                   & 66                  & 67                \\
			\NMIACE                      & \Bull{B} &  \Stop  &   \NoMg   & 61                                   & 60                  & 60                \\
			\OneMSevCE                       & \Bull{B} &  \Stop  &   \NoMg   & 60                                   & 61                  & 61                \\
			\BzCN                        & \Bull{B} &  \Stop  &   \NoMg   & 60                                   & 59                  & 59                \\
			\NAICE                         & \Bull{B} &  \Stop  &   \NoMg   & 60                                   & 59                  & 59                \\ \midrule
			Avg Technology              &          &  \Mut   &  $\star$  & 78                                   & 74                  & 75                \\
			                            &          &  \Stop  &  $\star$  & 66                                   & 65                  & 65                \\
			Avg -/+ Mg$^{2+}$           &          & $\star$ &   \NoMg   & 67                                   & 64                  & 65                \\
			                            &          & $\star$ &    \Mg    & 73                                   & 72                  & 71.5               \\ \midrule
			{\bfseries Average overall} &          & $\star$ &  $\star$  & 70.5                                 & 68                  & 68.5               \\ \bottomrule
			                            &          &         &           &
		\end{tabular}
		\end{adjustbox}\\[1em]}

@@

A superficial inspection of the resulting MCCs appears to confirm the conclusions of Section Bruno~\ref{sec:cordero}. Indeed, considering two conditions leads to predictions whose quality fall between the worst and the best one, while being generally close to the average. More precisely, the average MCC value of pair-informed predictions remains at 70.7\%, improving only by 0.2\% over the average MCC of mono-probing predictions. Pair-informed predictions are marginally more reliable, with the median MCC increases to 72.5\% MCC from 70.5\% for mono-probing predictions. However, outliers strongly impact this picture, and ignoring the conditions in cluster \Bull{B} increases the average MCC to 75.1\%, while unsurprisingly decreasing the standard deviation (8.6\% $\to$ 6.5\%).

@@

\paragraph{Considering diverse conditions improves and stabilizes the quality of predictions.} 
Having established the redundant nature of similar conditions, and their limited contribution to downstream modeling, we turn to an analysis of conditions across clusters.  We run \OurTool{} the quality of predictions obtained by choosing a single condition in each of the 8 clusters.\bsi{il manque des trucs dans cette phrase (celle qui commence par we run), non?} We consider any possible combination, except for the outlier cluster~\Bull{B}, for which we only consider \NAICE as a representative.

@@

The average MCC again increases in the presence of 3 mutants, to reach an average of 69.3\% MCC. For 10 mutants, the average MCC and overall distribution remains highly similar to that obtained by considering three mutants, suggesting peeking performances. These results support a notion of complementarity between the probing data produced across reasonably similar mutants, leading to a gradual decrease of the noise in the data. However, the bimodal nature of the distribution also suggests the existence of at least two conformations that are both supported by the WT and the mutant profiles.
\bsi{il y a une conformation alternative connue. Tu pourrais me donner un model provenant des 2 clusters de MCC que je regarde si pas hasard c'est ca?}

@@

\begin{figure}
	{\centering\includegraphics[width=1.2\linewidth]{graphs/boxplotreproducibility.png}\\}

>>>>

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al} dataset, further described. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics, for the sake of comparison with previous studies~\cite{Spasic2017}. As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data.

@@

This observed discrepancy is somewhat surprising, given that \OurTool and \Software{RSample} share the same sampling/clustering based modeling. However, the two methods differ substantially whenever a single conformation is detected, in which case \Software{RSample} switches to a simple MEA prediction, whereas \OurTool{} always proceeds with an iterative clustering. This allows \OurTool{} to consider a larger number of clusters, which we believe helps eliminating structural outliers, leading to the election of a better final centroid structure.

@@

	\caption{Model of the Lariat capping ribozyme from {\itshape Didymium iridis} (A -- PDB 4P8Z:A by Meyer\etal\cite{Meyer2014}), secondary structure, as annotated by \Software{DSSR} (B -- pseudoknot ommitted) and \OurTool first-ranking prediction (C.), obtained from the integration of 8 maximally-diverging probing conditions (\NMIAMg, \OneMSevILUThreeMg, \NAIMg, \OneMSevCE, \CMCTMg,\OneMSevILUThree, and \OneMSevMgCE).\label{fig:didy}}
\end{figure*}
As benchmark RNA we used the lariat capping ribozyme from \textit{Didymium Iridis} (DiLCrz) which 3D structure has been modeled at 3,5 A from X-ray diffraction data~\cite{Meyer2014}. This RNA contains a large diversity of interactions, including helical domains, tertiary pairings, a kissing loop and a pseudoknot (Figure~\ref{fig:didy}).
It was probed as described in material and methods in 16 different conditions using different SHAPE and classical reagents, in presence/absence of Mg$^\text{2+}$ ions, using either the standard premature stop or the newly developed mutation induction technologies to map the modification sites. These experimental results were then used as input of \OurTool{}, either individually or combined,  for secondary structure prediction.

@@

A visual inspection of Figure~\ref{fig:PCA} suggests the presence of 8 clusters. In order to objectively build groups of compatible conditions, we performed a k-mean clustering using \Software{scikit-learn}, setting k to 8, and obtained the clusters highlighted in Figures~\ref{fig:biclustering} and \ref{fig:PCA}.  We obtain the following clusters:
\begin{bulletlist}%\setlength{\itemsep}{0pt}\setlength{\parskip}{.1em}	
\item \BullLab{A}: \OneMSevILU, \NMIA, and \NMIAMg;
\item \BullLab{B}: \OneMSevCE, \NMIACE, \BzCN, and \NAICE;
\item \BullLab{C}: \OneMSevILUThreeMg and \OneMSevILUMg;
\item \BullLab{D}: \OneMSevILUThree;
\item \BullLab{E}: \NMIAMgCE and \OneMSevMgCE;
\item \BullLab{F}: \DMSMg and probing-free;
\item \BullLab{G}: \CMCTMg and \NAIMg;
\item \BullLab{H}: \BzCNMg.
\end{bulletlist}
The predicted clusters are generally consistent with the ordering resulting from a spectral biclustering, implemented in \Software{scikit-learn} and executed with default parameters, as illustrated by Figure~\ref{fig:glycine_example}.

@@

\begin{table}
	\Heatset{min=60,   
		max=83,   
		max colour=Aquamarine3, % colour at maximum
		min colour=BadCol,      % colour at minimum
		Min colour=BadCol, % colour for values below min
		Max colour=Aquamarine3   % colour for values above max
	}
	\newcommand{\Mut}{Mut}
	\newcommand{\Stop}{Stop}
	\newcommand{\Mg}{$\bullet$}
	\newcommand{\NoMg}{$\circ$}
	\newcommand{\M}[1]{\multicolumn{1}{l}{#1}}

	{\centering
			\begin{adjustbox}{max width=\linewidth}
		\begin{tabular}{@{}lcccHHH}
			\toprule
			                            &          &         &           & \multicolumn{3}{@{}c@{}}{MCC\%}              \\
			Condition                   &  Clust.  &  Tech.  & Mg$^{2+}$ & \M{\relsize{-1}\OurTool} & \M{MFE} & \M{MEA} \\ \midrule
			\OneMSevILUMg               & \Bull{C} &  \Mut   &    \Mg    & 85                       & 82      & 82      \\
			\OneMSevILU                 & \Bull{A} &  \Mut   &   \NoMg   & 84                       & 82      & 84      \\
			\NMIAMg                     & \Bull{A} &  \Mut   &    \Mg    & 81                       & 80      & 80      \\
			\NMIA                       & \Bull{A} &  \Mut   &   \NoMg   & 80                       & 69      & 69      \\
			\OneMSevILUThreeMg          & \Bull{C} &  \Mut   &    \Mg    & 74                       & 75      & 75      \\
			\NMIAMgCE                   & \Bull{E} &  \Stop  &    \Mg    & 73                       & 72      & 73      \\
			\NAIMg                      & \Bull{G} &  \Stop  &    \Mg    & 73                       & 68      & 70      \\
			\BzCNMg                     & \Bull{H} &  \Stop  &    \Mg    & 71                       & 68      & 54      \\
			\OneMSevMgCE                & \Bull{E} &  \Stop  &    \Mg    & 70                       & 69      & 70      \\
			\CMCTMg                     & \Bull{G} &  \Stop  &    \Mg    & 70                       & 68      & 73      \\
			\OneMSevILUThree            & \Bull{D} &  \Mut   &   \NoMg   & 64                       & 56      & 61      \\
			\DMSMg                      & \Bull{F} &  \Stop  &    \Mg    & 63                       & 66      & 67      \\
			\NMIACE                     & \Bull{B} &  \Stop  &   \NoMg   & 61                       & 60      & 60      \\
			\OneMSevCE                  & \Bull{B} &  \Stop  &   \NoMg   & 60                       & 61      & 61      \\
			\BzCN                       & \Bull{B} &  \Stop  &   \NoMg   & 60                       & 59      & 59      \\
			\NAICE                      & \Bull{B} &  \Stop  &   \NoMg   & 60                       & 59      & 59      \\ \midrule
			Avg Technology              &          &  \Mut   &  $\star$  & 78                       & 74      & 75      \\
			                            &          &  \Stop  &  $\star$  & 66                       & 65      & 65      \\
			Avg -/+ Mg$^{2+}$           &          & $\star$ &   \NoMg   & 67                       & 64      & 65      \\
			                            &          & $\star$ &    \Mg    & 73                       & 72      & 71.5    \\ \midrule
			{\bfseries Average overall} &          & $\star$ &  $\star$  & 70.5                     & 68      & 68.5    \\ \bottomrule
			                            &          &         &           &                          &
		\end{tabular}
		\end{adjustbox}\\[1em]}

@@

A superficial inspection of the resulting MCCs appears to confirm the conclusions reached on the Cordero~\emph{et al}~\cite{Cordero2012} dataset. Indeed, considering two conditions leads to predictions whose quality fall between the worst and the best one, while being generally close to the average. More precisely, the average MCC value of pair-informed predictions remains at 70.7\%, improving only by 0.2\% over the average MCC of mono-probing predictions. Pair-informed predictions are marginally more reliable, with the median MCC increases to 72.5\% MCC from 70.5\% for mono-probing predictions. However, outliers strongly impact this picture, and ignoring the conditions in cluster \Bull{B} increases the average MCC to 75.1\%, while unsurprisingly decreasing the standard deviation (8.6\% $\to$ 6.5\%).

@@

\paragraph{Considering diverse conditions improves and stabilizes the quality of predictions.} 
Having established the redundant nature of similar conditions, and their limited contribution to downstream modeling, we turn to an analysis of conditions across clusters.  We run \OurTool{} to assess the quality of predictions informed by condition chosen across the 8 clusters.We consider any possible combination, except for the outlier cluster~\Bull{B}, for which we only consider \NAICE as a representative.

@@

The average MCC again increases in the presence of 3 mutants, to reach an average of 69.3\% MCC. For 10 mutants, the average MCC and overall distribution remains highly similar to that obtained by considering three mutants, suggesting peeking performances. These results support a notion of complementarity between the probing data produced across reasonably similar mutants, leading to a gradual decrease of the noise in the data. However, the bimodal nature of the distribution also suggests the existence of at least two conformations that are both supported by the WT and the mutant profiles.

@@

\begin{figure}
	{\centering\includegraphics[width=1\linewidth]{graphs/boxplotreproducibility.png}\\}
